Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-09-25 manuscript/climate_sensitivity_SI.Rmd
Updated for New Phyt guidelines #148

\newpage
![**Figure S2. Map of ForestGEO plot showing topographic wetness index and location of cored trees.** Scale units are in meters](tables_figures/publication/figureS2_location_cored_trees.png){width=500px}

>>>>

\newpage
![**Figure S2. Map of ForestGEO plot showing topographic wetness index (color scale) and location of cored trees.** Scale units are in meters](tables_figures/publication/figureS2_location_cored_trees.png){width=500px}

### 2020-09-25 manuscript/climate_sensitivity_draft.Rmd
Updated for New Phyt guidelines #148

Moreover, it has yet to be resolved which of several potential underlying mechanisms most strongly shape these trends in drought response. 
First, tree height itself may be a primary driver. 
Taller trees face the biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018]. 
Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018].
Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts [@phillips_reliance_2003; @scholz_hydraulic_2011].
Taller trees have wider conduits in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019] and throughout the conductive systems of angiosperms [@zach_vessel_2010; @olson_universal_2014; @olson_plant_2018], which help maintain constant the resistance that would otherwise increase as trees grow taller.
Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].

@@

Tree heights ($H$) were measured by several researchers for a variety of purposes between 2012 and 2019 (n=1,518 trees). Methods included direct measurements using a collapsible measurement rod on small trees [@neon_national_2018] or a tape measure on recently fallen trees (this study); geometric calculations using clinometer and tape measure [@stovall_assessing_2018] or digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018]; and ground-based LiDAR [@stovall_terrestrial_2018]. Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other. 
Species-specific height allometries were developed using log-log regression ($ln[H] \sim ln[DBH]$; Table S3). 
For species with insufficient height data to create reliable species-specific allometries (n=2, JUNI and FRAM), heights were calculated from an equation developed by combining the height measurements across all species. 
We then used these allometries to estimate $H$ for each drought year, $Y$, based on reconstructed $DBH_Y$ (Fig. **S3**).

@@

Trees in drier microsites showed greater growth declines during drought; *i.e.*, $Rt$ had a significantly negative response to $ln[TWI]$ across all drought years combined, and in 1977 and 1999 individually (Fig. **4**, Table S8-S9). 
The $ln[H] *ln[TWI]$ interaction was never significant, and had a positive sign in any top $Rt$ models in which it appeared (Tables 1, S8-S9), rejecting the hypothesis that smaller trees (presumably with smaller rooting volume) are more susceptible to drought in microenvironments with a deeper water table. 
In contrast, $ln[TWI]$ did not appear in any of the best models for $Rc$ or $Rs$ (combined of for individual years), except in interaction with $ln[H]$ (Fig. 4, Tables S10-S11).
Negative $ln[H] *ln[TWI]$ interactions appeared in the best models for both $Rc$ and $Rs$ for all years combined, as well as in one individual year for each (1966 for $Rc$, 1977 for $Rs$). 
This implies a non-significant tendency for small trees to have greater recovery and resilience in wetter microhabitats, but for large trees to have greater recovery and resilience in dry microhabitats.

@@

Species, as a factor in ANOVA, had significant (p<0.05) influence on all traits (wood density, $LMA$, $PLA_{dry}$, and $\pi_{tlp}$), with more significant pairwise differences for wood density and $PLA_{dry}$ than for $LMA$ and $\pi_{tlp}$ (Table 2, Fig. **S4**).
Drought tolerance also varied across species, overall and in each drought year (Figs. **3, S7**). 
Species with overall lowest and highest $Rt$ and $Rs$ were, respectively, *Liriodendron tulipifera* (mean $Rt$ = 0.66, mean $Rs$ = 1.04) and *Fagus grandifolia* (mean $Rt$ = 0.99; mean $Rs$ = 1.65).
These two species--notably the only two diffuse-porous species in our study--differed significantly from one another in $Rt$ and $Rs$ in each drought year (Fig. **3**).

Wood density, $LMA$, and xylem porosity were all poor predictors of drought tolerance (Tables 1, S4-S5). 
Wood density and $LMA$ were never significantly associated with $Rt$, $Rc$, or $Rs$ in the single-variable tests and were therefore excluded from the full models. 
Xylem porosity had no significant influence on $Rt$ or $Rs$ in models for all droughts combined (Tables S4, S7). 
In contrast, $Rc$ was significantly higher in diffuse- and semi-ring porous species than in ring-porous species (Table S6, Fig. 3).

@@

The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years [Fig. **1a**, @helcoski_growing_2019] and had substantial but short-lived impacts on tree growth (Fig. **1**). 
These droughts were classified as severe ($PDSI$ < -3.0; 1977) or extreme ($PDSI$ < -4.0; 1966, 1999) at our site and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019], but were modest compared to the so-called “megadroughts” that have triggered massive tree die-off in other regions [e.g., @allen_global_2010; @stovall_tree_2019; @clark_impacts_2016]. 
Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig. **S1**), severely stressed a larger portion of trees (Fig. **1b**). 
The tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019]. 
Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion (e.g., short understory trees, species with drought resistant traits) had increased growth (Figs. **1b, 4**), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]. 
Growth rebounded strongly following the droughts, on average exceeding pre-drought growth rates (Fig. 1), particularly for shorter trees and species with drought-tolerant traits (Figs. **3-4**).
It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$, $Rc$, and $Rs$: 11-18% for all droughts combined, and 13-30% for individual droughts (Tables S8-S11).

Consistent with studies in other forests worldwide [@bennett_larger_2015], taller trees in this forest exhibited lower drought resistance--and also recovery and resilience--when compared to smaller trees.
Mechanistically, this is consistent with, and reinforces, previous findings that it is impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019]. 
Taller trees also face dramatically different microenvironments (Fig. **2**). 
They are exposed to higher wind speeds and lower humidity (Fig. **2a-b**), resulting in higher evaporative demand. 
Unlike other temperate forests where modestly cooler understory conditions have been documented [@zellweger_seasonal_2019], particularly under drier conditions [@davis_microclimatic_2019], we observed no significant variation in air temperatures across the vertical profile (Fig. **2c**). 
More critically for tree physiology, leaf temperatures can become significantly elevated over air temperature under conditions of high solar radiation and low stomatal conductance [@campbell_introduction_1998; @rey-sanchez_spatial_2016]. 
Under drought, when direct solar radiation tends to be higher (because of less cloud cover) and less water is available for evaporative cooling of the leaves, trees with sun-exposed crowns may not be able to simultaneously maintain leaf temperatures below damaging extremes and avoid drought-induced embolism. 
Indeed, previous studies have shown lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019].
Unfortunately, collinearity between height and crown exposure in this study (Fig. **2d**) makes it impossible to confidently partition causality. 
Additional research comparing drought responses of early successional and mature forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.

@@

The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species (Fig. **3**) and associated traits at a single site [see also @elliott_forest_2015]. 
Our study reinforced current understanding (see Introduction) that wood density and $LMA$ are not reliably linked to drought tolerance (Table 1).
Contrary to several previous studies in temperate deciduous forests [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], we did not find an association between xylem porosity and drought resistance or resilience, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig. **3**). 
While the low $Rt$ of *L. tulipifera* is consistent with other studies [@elliott_forest_2015], the high $Rt$ of *F. grandifolia* contrasts with studies identifying diffuse porous species in general [@elliott_forest_2015; @kannenberg_linking_2019], and the genus *Fagus* in particular [@friedrichs_species-specific_2009], as drought sensitive.
There are two potential explanations for this discrepancy.
First, other traits can and do override the influence of xylem porosity on drought resistance. 
Ring-porous species are restricted mainly to temperate deciduous forests, while highly drought-tolerant diffuse-porous species exist in other biomes [@wheeler_variations_2007]. 
*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to its drought tolerance (Fig. **4**; see discussion below), in concordance with studies identifying *Fagus* species as intermediate in drought tolerance [@vitasse_contrasting_2019; @pretzsch_drought_2018]. 
A second explanation of why *F. grandifolia* trees at this particular site had higher $Rt$ and $Rs$ is that the sampled individuals, reflective of the population within the plot, are generally shorter and in less-dominant canopy positions compared to most other species (Fig. **S4**).
The species, which is highly shade-tolerant, also has deep crowns [@andersonteixeira_size-related_2015], implying that a lower proportion of leaves would be affected by harsher microclimatic conditions at the top of the canopy under drought (Fig. **2**).
Thus, the high $Rt$ and $Rs$ of the sampled *F. grandifolia* population can be explained by a combination of fairly drought-resistant leaf traits, shorter stature, and a buffered microenvironment.

>>>>

Moreover, it has yet to be resolved which of several potential underlying mechanisms most strongly shape these trends in drought response. 
First, tree height itself may be a primary driver. 
Taller trees face the biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018]. 
Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018].
Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts [@phillips_reliance_2003; @scholz_hydraulic_2011].
Taller trees have wider conduits in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019] and throughout the conductive systems of angiosperms [@zach_vessel_2010; @olson_universal_2014; @olson_plant_2018], which help maintain constant the resistance that would otherwise increase as trees grow taller.
Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].

@@

Tree heights ($H$) were measured by several researchers for a variety of purposes between 2012 and 2019 (n=1,518 trees). Methods included direct measurements using a collapsible measurement rod on small trees [@neon_national_2018] or a tape measure on recently fallen trees (this study); geometric calculations using clinometer and tape measure [@stovall_assessing_2018] or digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018]; and ground-based LiDAR [@stovall_terrestrial_2018]. Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other. 
Species-specific height allometries were developed using log-log regression ($ln[H] \sim ln[DBH]$; Table S3). 
For species with insufficient height data to create reliable species-specific allometries (n=2, *Juglans nigra* (JUNI) and *Fraxinus americana* (FRAM)), heights were calculated from an equation developed by combining the height measurements across all species. 
We then used these allometries to estimate $H$ for each drought year, $Y$, based on reconstructed $DBH_Y$ (Fig. **S3**).

@@

Trees in drier microsites showed greater growth declines during drought; *i.e.*, $Rt$ had a significantly negative response to $ln[TWI]$ across all drought years combined, and in 1977 and 1999 individually (Fig. **4**, Table S8-S9). 
The $ln[H] *ln[TWI]$ interaction was never significant, and had a positive sign in any top $Rt$ models in which it appeared (Tables 1, S8-S9), rejecting the hypothesis that smaller trees (presumably with smaller rooting volume) are more susceptible to drought in microenvironments with a deeper water table. 
In contrast, $ln[TWI]$ did not appear in any of the best models for $Rc$ or $Rs$ (combined of for individual years), except in interaction with $ln[H]$ (Fig. **4**, Tables S10-S11).
Negative $ln[H] *ln[TWI]$ interactions appeared in the best models for both $Rc$ and $Rs$ for all years combined, as well as in one individual year for each (1966 for $Rc$, 1977 for $Rs$). 
This implies a non-significant tendency for small trees to have greater recovery and resilience in wetter microhabitats, but for large trees to have greater recovery and resilience in dry microhabitats.

@@

Species, as a factor in ANOVA, had significant (p<0.05) influence on all traits (wood density, $LMA$, $PLA_{dry}$, and $\pi_{tlp}$), with more significant pairwise differences for wood density and $PLA_{dry}$ than for $LMA$ and $\pi_{tlp}$ (Table 2, Fig. **S4**).
Drought tolerance also varied across species, overall and in each drought year (Figs. **3, S7**). 
Species with overall lowest and highest $Rt$ and $Rs$ were, respectively, *L. tulipifera* (mean $Rt$ = 0.66, mean $Rs$ = 1.04) and *Fagus grandifolia* (mean $Rt$ = 0.99; mean $Rs$ = 1.65).
These two species--notably the only two diffuse-porous species in our study--differed significantly from one another in $Rt$ and $Rs$ in each drought year (Fig. **3**).

Wood density, $LMA$, and xylem porosity were all poor predictors of drought tolerance (Tables 1, S4-S5). 
Wood density and $LMA$ were never significantly associated with $Rt$, $Rc$, or $Rs$ in the single-variable tests and were therefore excluded from the full models. 
Xylem porosity had no significant influence on $Rt$ or $Rs$ in models for all droughts combined (Tables S4, S7). 
In contrast, $Rc$ was significantly higher in diffuse- and semi-ring porous species than in ring-porous species (Table S6, Fig. **3**).

@@

The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years [Fig. **1a**, @helcoski_growing_2019] and had substantial but short-lived impacts on tree growth (Fig. **1**). 
These droughts were classified as severe ($PDSI$ < -3.0; 1977) or extreme ($PDSI$ < -4.0; 1966, 1999) at our site and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019], but were modest compared to the so-called “megadroughts” that have triggered massive tree die-off in other regions [e.g., @allen_global_2010; @stovall_tree_2019; @clark_impacts_2016]. 
Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig. **S1**), severely stressed a larger portion of trees (Fig. **1b**). 
The tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019]. 
Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion (e.g., short understory trees, species with drought resistant traits) had increased growth (Figs. **1b, 4**), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]. 
Growth rebounded strongly following the droughts, on average exceeding pre-drought growth rates (Fig. **1**), particularly for shorter trees and species with drought-tolerant traits (Figs. **3-4**).
It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$, $Rc$, and $Rs$: 11-18% for all droughts combined, and 13-30% for individual droughts (Tables S8-S11).

Consistent with studies in other forests worldwide [@bennett_larger_2015], taller trees in this forest exhibited lower drought resistance--and also recovery and resilience--when compared to smaller trees.
Mechanistically, this is consistent with, and reinforces, previous findings that it is impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019]. 
Taller trees also face dramatically different microenvironments (Fig. **2**). 
They are exposed to higher wind speeds and lower humidity (Fig. **2a,b**), resulting in higher evaporative demand. 
Unlike other temperate forests where modestly cooler understory conditions have been documented [@zellweger_seasonal_2019], particularly under drier conditions [@davis_microclimatic_2019], we observed no significant variation in air temperatures across the vertical profile (Fig. **2c**). 
More critically for tree physiology, leaf temperatures can become significantly elevated over air temperature under conditions of high solar radiation and low stomatal conductance [@campbell_introduction_1998; @rey-sanchez_spatial_2016]. 
Under drought, when direct solar radiation tends to be higher (because of less cloud cover) and less water is available for evaporative cooling of the leaves, trees with sun-exposed crowns may not be able to simultaneously maintain leaf temperatures below damaging extremes and avoid drought-induced embolism. 
Indeed, previous studies have shown lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019].
Unfortunately, collinearity between height and crown exposure in this study (Fig. **2d**) makes it impossible to confidently partition causality. 
Additional research comparing drought responses of early successional and mature forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.

@@

The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species (Fig. **3**) and associated traits at a single site [see also @elliott_forest_2015]. 
Our study reinforced current understanding (see Introduction) that wood density and $LMA$ are not reliably linked to drought tolerance (Table 1).
Contrary to several previous studies in temperate deciduous forests [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], we did not find an association between xylem porosity and drought resistance or resilience, as the two diffuse-porous species, *L.* and *F. grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig. **3**). 
While the low $Rt$ of *L. tulipifera* is consistent with other studies [@elliott_forest_2015], the high $Rt$ of *F. grandifolia* contrasts with studies identifying diffuse porous species in general [@elliott_forest_2015; @kannenberg_linking_2019], and the genus *Fagus* in particular [@friedrichs_species-specific_2009], as drought sensitive.
There are two potential explanations for this discrepancy.
First, other traits can and do override the influence of xylem porosity on drought resistance. 
Ring-porous species are restricted mainly to temperate deciduous forests, while highly drought-tolerant diffuse-porous species exist in other biomes [@wheeler_variations_2007]. 
*F. grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to its drought tolerance (Fig. **4**; see discussion below), in concordance with studies identifying *Fagus* species as intermediate in drought tolerance [@vitasse_contrasting_2019; @pretzsch_drought_2018]. 
A second explanation of why *F. grandifolia* trees at this particular site had higher $Rt$ and $Rs$ is that the sampled individuals, reflective of the population within the plot, are generally shorter and in less-dominant canopy positions compared to most other species (Fig. **S4**).
The species, which is highly shade-tolerant, also has deep crowns [@andersonteixeira_size-related_2015], implying that a lower proportion of leaves would be affected by harsher microclimatic conditions at the top of the canopy under drought (Fig. **2**).
Thus, the high $Rt$ and $Rs$ of the sampled *F. grandifolia* population can be explained by a combination of fairly drought-resistant leaf traits, shorter stature, and a buffered microenvironment.

### 2020-09-25 manuscript/tables&figures.Rmd
moved content from results to fig caption
Updated for New Phyt guidelines #148

```{r, Table 2, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
table2 <- read.csv("tables_figures/publication/table2_species_table_hand_formatted.csv", stringsAsFactors = FALSE, check.names = FALSE)
table2[0,3] <- paste0(table2[0,3], footnote_marker_symbol(1))
table2[8,8] <- paste0(table2[8,8], footnote_marker_symbol(2))
kable(table2, format = "latex", booktabs=TRUE, caption = "Table 2. Overview of analyzed species, listed in order of their relative contributions to woody stem productivity ($ANPP_{stem}$) in the plot, along with numbers and sizes sampled, and species traits. Variable abbreviations are as in Table 3. DBH measurements are from the most recent ForestGEO census in 2018 (live trees) or tree mortality censuses in 2016 and 2017 (trees cored dead).", escape=FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," ", "contemporary DBH (cm)" = 2, "species traits (mean +/- se)" = 5)) %>%
  #kable_styling(bootstrap_options = "striped", full_width = FALSE)
kableExtra::footnote(symbol = c("Numbers cored live versus dead are given in Table S1 of Helcoski et al. (2019).", "Semi-ring porosity is intermediate between ring and diffuse. We group it with diffuse-porous species for more even division of species between categories."), footnote_as_chunk = FALSE)
```

@@

**Figure 1. Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide growth resistance,$Rt$ (b), and resilience, $Rs$ (c).** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes). PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al. 2014). Focal droughts are indicated by dashed lines, and shading indicates the pre- and post- drought periods used in calculations of the resistance metric. Figure modified from Helcoski *et al.* (2019). Density plots (b-c) show the distribution of $Rt$ and $Rs$ values for each drought. See Fig. S6 for parallel plot for recovery ($Rc$).

@@

**Figure 3. Drought resistance, $Rt$ (a), and resilience, $Rs$ (b), across species for the three focal droughts.** Species codes are given in Table 2. See Fig. S7 for parallel plot for recovery ($Rc$).

@@

![**Figure 1. Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide growth resistance,$Rt$ (b), and resilience, $Rs$ (c).** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes). PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al. 2014). Focal droughts are indicated by dashed lines, and shading indicates the pre- and post- drought periods used in calculations of the resistance metric. Figure modified from Helcoski *et al.* (2019). Density plots (b-c) show the distribution of $Rt$ and $Rs$ values for each drought. See Fig. S6 for parallel plot for recovery ($Rc$). ](tables_figures/publication/Figure1.png)

>>>>

```{r, Table 2, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
table2 <- read.csv("tables_figures/publication/table2_species_table_hand_formatted.csv", stringsAsFactors = FALSE, check.names = FALSE)
table2[0,3] <- paste0(table2[0,3], footnote_marker_symbol(1))
table2[8,8] <- paste0(table2[8,8], footnote_marker_symbol(2))
kable(table2, format = "latex", booktabs=TRUE, caption = "Table 2. Overview of analyzed species, listed in order of their relative contributions to woody stem productivity ($ANPP_{stem}$) in the plot, along with numbers and sizes sampled, and species traits.", escape=FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," ", "contemporary DBH (cm)" = 2, "species traits (mean +/- se)" = 5)) %>%
  #kable_styling(bootstrap_options = "striped", full_width = FALSE)
kableExtra::footnote(symbol = c("Numbers cored live versus dead are given in Table S1 of Helcoski et al. (2019).", "Semi-ring porosity is intermediate between ring and diffuse. We group it with diffuse-porous species for more even division of species between categories."), footnote_as_chunk = FALSE)
```
Variable abbreviations are as in Table 3. DBH measurements are from the most recent ForestGEO census in 2018 (live trees) or tree mortality censuses in 2016 and 2017 (trees cored dead).

@@

**Figure 1. Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide growth resistance, $Rt$ (b), and resilience, $Rs$ (c).** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes). PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al. 2014). Focal droughts are indicated by dashed lines, and shading indicates the pre- and post- drought periods used in calculations of the resistance metric. Figure modified from Helcoski *et al.* (2019). Density plots (b-c) show the distribution of $Rt$ and $Rs$ values for each drought. See Fig. S6 for parallel plot for recovery ($Rc$).

@@

**Figure 3. Drought resistance, $Rt$ (a), and resilience, $Rs$ (b), across species for the three focal droughts.** Species codes are given in Table 2. See Fig. S7 for parallel plot for recovery ($Rc$). Analysis conducted using *agricolae* package [@R-agricolae].

@@

![**Figure 1. Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide growth resistance, $Rt$ (b), and resilience, $Rs$ (c).** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes). PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al. 2014). Focal droughts are indicated by dashed lines, and shading indicates the pre- and post- drought periods used in calculations of the resistance metric. Figure modified from Helcoski *et al.* (2019). Density plots (b-c) show the distribution of $Rt$ and $Rs$ values for each drought. See Fig. S6 for parallel plot for recovery ($Rc$). ](tables_figures/publication/Figure1.png)
